---
title: "Untitled"
author: "Mamadou"
date: "2023-11-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
setwd(F:/diabetes_health_indicators.csv)
## Initial Setup: Required Libraries and Working Directory
```{r message=F}
library(tidyverse)
library(sf)
library(plotly)
library(caret)
library(likert)
library(grid)
library(gridExtra)
library(ggpubr)
```


## Data Wrangling

## Load and Inspect the Data

```{r}
# load the data


data <- read.csv("diabetes_health_indicators.csv")
```

Removing unwanted variables

```{r}
data_select <- data %>%
        select(-c(7,14,18))
```

let's check the structure of the dataset
```{r}
str(data_select)
```
We can see that all data are numeric. However some are actually categorical ones so we need to convert them to factor data type.

```{r}
data_select$Diabetes_binary <- as.factor(data_select$Diabetes_binary)
data_select$HighBP <- as.factor(data_select$HighBP)
data_select$CholCheck <- as.factor(data_select$CholCheck)
data_select$Smoker <- as.factor(data_select$Smoker)
data_select$HeartDiseaseorAttack <- as.factor(data_select$HeartDiseaseorAttack)
data_select$PhysActivity <- as.factor(data_select$PhysActivity)
data_select$Fruits <- as.factor(data_select$Fruits)
data_select$Veggies <- as.factor(data_select$Veggies)
data_select$HvyAlcoholConsump <- as.factor(data_select$HvyAlcoholConsump)
data_select$AnyHealthcare <- as.factor(data_select$AnyHealthcare)
data_select$Sex <- as.factor(data_select$Sex)
data_select$HighChol <-as.factor(data_select$HighChol)
```

## Data Subset
Since we have 70692 observations, we need to subset the data a lower dataset and we opt to move to 2000 observation and 10 variables.
```{r}
#diabetes_subset <- data_select %>%
  #select(c("HighBP", "HighChol", "BMI", "Smoker", "HeartDiseaseorAttack", "AnyHealthcare", "Sex", "Age", "Education", "Income", "Diabetes_binary"))
```

## Create Index for even split of target variable
```{r}
#set.seed(123)
diabetes_sample <- data_select %>%
  group_by(Diabetes_binary) %>%
  slice_sample(n = 1000) %>%
  ungroup() #%>%
  #slice_sample(n = 2000)
#summary(diabetes_sample$Diabetes_binary)
```

## Create Training Data
```{r}
#set.seed(123)
inTraining <- createDataPartition(diabetes_sample$Diabetes_binary, p=0.75, list=FALSE)
training <- diabetes_sample[inTraining,]
validation <- diabetes_sample[-inTraining,]
validation
```

## Establish training parameters
```{r}
control <- trainControl(method="cv", number=10)
metric <- "Accuracy"
```

## Train the model
```{r}
# linear algorithm (Linear Discriminant Analysis)
#set.seed(123)
fit.lda <- train(Diabetes_binary~., data=training, method="lda", metric=metric, trControl=control)
predictionsLDA <- predict(fit.lda, validation)
predictionsLDA

cmLDA <- confusionMatrix(predictionsLDA, as.factor(validation$Diabetes_binary))
cmLDA

# Classification and Regression Trees
#set.seed(123)
fit.cart <- train(Diabetes_binary~., data=training, method="rpart", metric=metric, trControl=control)
predictionsCRT <- predict(fit.cart, validation)

cmCRT <- confusionMatrix(predictionsCRT, as.factor(validation$Diabetes_binary))
cmCRT

# k-Nearest Neighbors
#set.seed(123)
fit.knn <- train(Diabetes_binary~., data=training, method="knn", metric=metric, trControl=control)
predictionsKNN <- predict(fit.knn, validation)

cmKNN <- confusionMatrix(predictionsKNN, as.factor(validation$Diabetes_binary))
cmKNN

# Support Vector Machines
set.seed(123)
fit.svm <- train(Diabetes_binary~., data=training, method="svmRadial", metric=metric, trControl=control)
predictionsSVM <- predict(fit.svm, validation)

cmSVM <- confusionMatrix(predictionsSVM, as.factor(validation$Diabetes_binary))
cmSVM

# Logistic Regression
#set.seed(123)
fit.glm <- train(Diabetes_binary~., data=training, method="glm", metric=metric, trControl=control)
predictionsLR <- predict(fit.glm, validation)

cmLR <- confusionMatrix(predictionsLR, as.factor(validation$Diabetes_binary))
cmLR
```


## evaluate the model
```{r}
results <- resamples(list(lda=fit.lda, cart=fit.cart, knn=fit.knn, svm=fit.svm, glm=fit.glm))
summary(results)

results_df <- as.data.frame(results)

results_tidy <- results_df %>% 
  pivot_longer(names_to = "Model", values_to = "Accuracy", -Resample) %>% 
  group_by(Model) %>% 
  summarise(Mean_Accuracy = mean(Accuracy))

mean_acc <- results_tidy %>% 
  ggplot(aes(x=fct_reorder(Model, Mean_Accuracy), y=Mean_Accuracy))+
  geom_bar(stat = "identity")+
  coord_flip()+
  xlab("Mean Accuracy")+
  ylab("Model")+
  theme(text = element_text(size = 20))

mean_acc
```

