---
title: "Untitled"
author: "Mamadou"
date: "2023-11-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
setwd(F:/diabetes_health_indicators.csv)
## Initial Setup: Required Libraries and Working Directory
```{r message=F}
library(tidyverse)
library(sf)
library(plotly)
library(caret)
library(likert)
library(grid)
library(gridExtra)
library(ggpubr)
```


## Data Wrangling

## Load and Inspect the Data

```{r}
# load the data


data <- read.csv("diabetes_health_indicators.csv")
```

Removing unwanted variables

```{r}
data_select <- data %>%
        select(-c(10:14,16:17,19))
```

let's check the structure of the dataset
```{r}
str(data_select)
```
We can see that all data are numeric. However some are actually categorical ones so we need to convert them to factor data type.

```{r}
data_select$Diabetes_binary <- as.factor(data_select$Diabetes_binary)
data_select$HighBP <- as.factor(data_select$HighBP)
data_select$CholCheck <- as.factor(data_select$CholCheck)
data_select$Smoker <- as.factor(data_select$Smoker)
data_select$HeartDiseaseorAttack <- as.factor(data_select$HeartDiseaseorAttack)
data_select$PhysActivity <- as.factor(data_select$PhysActivity)
data_select$HighChol <-as.factor(data_select$HighChol)
data_select$Stroke <-as.factor(data_select$Stroke)
data_select$GenHlth <-as.factor(data_select$GenHlth)
data_select$DiffWalk <-as.factor(data_select$DiffWalk)
data_select$Education <-as.factor(data_select$Education)
data_select$Income <-as.factor(data_select$Income)
```

## Data Subset
Since we have 70692 observations, we need to subset the data a lower dataset and we opt to move to 2000 observation and 10 variables.

```{r}
#Create Index for even split of target variable
diabetes_sample <- data_select %>%
  group_by(Diabetes_binary) %>%
  slice_sample(n = 5000) %>%
  ungroup() 
```

## Z transformation
```{r}
diabetes_scaled <- diabetes_sample %>% 
 mutate(across(.cols = c(5,12), ~ as.vector(scale(.)), .names = "scaled_{.col}")) %>%
  select(c(1:4, 6:11,13:16))
View(df_scaled)
```


## Create Training Data
```{r}
#set.seed(123)
inTraining <- createDataPartition(diabetes_scaled$Diabetes_binary, p=0.75, list=FALSE)
training <- diabetes_scaled[inTraining,]
validation <- diabetes_scaled[-inTraining,]
validation
```

## Establish training parameters
```{r}
control <- trainControl(method="cv", number=10)
metric <- "Accuracy"
```

## Train the model
```{r}
# linear algorithm (Linear Discriminant Analysis)
fit.lda <- train(Diabetes_binary~., data=training, method="lda", metric=metric, trControl=control)
predictionsLDA <- predict(fit.lda, validation)
predictionsLDA

cmLDA <- confusionMatrix(predictionsLDA, as.factor(validation$Diabetes_binary))
cmLDA
```

```{r}
# Classification and Regression Trees
fit.cart <- train(Diabetes_binary~., data=training, method="rpart", metric=metric, trControl=control)
predictionsCRT <- predict(fit.cart, validation)

cmCRT <- confusionMatrix(predictionsCRT, as.factor(validation$Diabetes_binary))
cmCRT
```

```{r}
# k-Nearest Neighbors
fit.knn <- train(Diabetes_binary~., data=training, method="knn", metric=metric, trControl=control)
predictionsKNN <- predict(fit.knn, validation)

cmKNN <- confusionMatrix(predictionsKNN, as.factor(validation$Diabetes_binary))
cmKNN
```

```{r}
# Support Vector Machines
fit.svm <- train(Diabetes_binary~., data=training, method="svmRadial", metric=metric, trControl=control)
predictionsSVM <- predict(fit.svm, validation)

cmSVM <- confusionMatrix(predictionsSVM, as.factor(validation$Diabetes_binary))
cmSVM
```

```{r}
# Logistic Regression
fit.glm <- train(Diabetes_binary~., data=training, method="glm", metric=metric, trControl=control)
predictionsLR <- predict(fit.glm, validation)

cmLR <- confusionMatrix(predictionsLR, as.factor(validation$Diabetes_binary))
cmLR
```


```{r}
# Random Forests
fit.rf <- train(Diabetes_binary ~., data=training, method="rf", metric=metric, trControl=control)
predictions.rf <- predict(fit.rf, validation)
metrics.rf <- postResample(predictions.rf, validation$Diabetes_binary)
metrics.rf

cmrf <- confusionMatrix(predictions.rf, as.factor(validation$Diabetes_binary))
cmrf
```


## evaluate the model
```{r}
results <- resamples(list(lda= fit.lda, cart=fit.cart, knn=fit.knn, svm=fit.svm, glm=fit.glm, rf=fit.rf))
summary(results)

results_df <- as.data.frame(results)

results_tidy <- results_df %>% 
  pivot_longer(names_to = "Model", values_to = "Accuracy", -Resample) %>% 
  group_by(Model) %>% 
  summarise(Mean_Accuracy = mean(Accuracy))

mean_acc <- results_tidy %>% 
  ggplot(aes(x=fct_reorder(Model, Mean_Accuracy), y=Mean_Accuracy))+
  geom_bar(stat = "identity")+
  coord_flip()+
  xlab("Mean Accuracy")+
  ylab("Model")+
  theme(text = element_text(size = 20))

mean_acc
```



```{r}
# Calculate variable importance for four of five models
importance1 <- varImp(fit.glm)
importance2 <- varImp(fit.rf)
importance3 <- varImp(fit.cart)

imp1 <- importance1$importance 
imp2 <- importance2$importance
imp3 <- importance3$importance

 p1 <- imp1 %>% 
   mutate(Predictor = rownames(imp1)) %>% 
   pivot_longer(names_to = "Diabetes_binary", values_to = "Importance", -Predictor) %>%
   ggplot(aes(x=Predictor, y=Importance))+
   geom_segment(aes(x=Predictor, xend=Predictor, y=0, yend=Importance), color="skyblue") +
   geom_point(color="blue", size=4, alpha=0.6) +
   theme_light() +
   coord_flip() +
   theme(
     panel.grid.major.y = element_blank(),
     panel.border = element_blank(),
     axis.ticks.y = element_blank())+
   ylab("Linear Regresssion")+
   xlab("")
 
 p2 <- imp2 %>% 
  mutate(Predictor = rownames(imp2)) %>% 
  pivot_longer(names_to = "Diabetes_binary", values_to = "Importance", -Predictor) %>%
  ggplot(aes(x=Predictor, y=Importance))+
  geom_segment(aes(x=Predictor, xend=Predictor, y=0, yend=Importance), color="skyblue") +
  geom_point(color="blue", size=4, alpha=0.6) +
  theme_light() +
  coord_flip() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank())+
  ylab("Random Forest")+
  xlab("")

 
 p3 <- imp3 %>%
   mutate(Predictor = rownames(imp3)) %>%
   pivot_longer(names_to = "Diabetes_binary", values_to = "Importance", -Predictor) %>%
   ggplot(aes(x=Predictor, y=Importance))+
   geom_segment(aes(x=Predictor, xend=Predictor, y=0, yend=Importance), color="skyblue") +
   geom_point(color="blue", size=4, alpha=0.6) +
   theme_light() +
   coord_flip() +
   theme(
     panel.grid.major.y = element_blank(),
     panel.border = element_blank(),
     axis.ticks.y = element_blank())+
   ylab("Cart")+
   xlab("")

average_importance <- rowMeans(cbind(imp1,imp2, imp3))
importance_df <- data.frame(Predictor = names(average_importance), Importance = average_importance)
importance_df <- importance_df[order(importance_df$Importance, decreasing = TRUE), ]
library(ggplot2)

ggplot(importance_df, aes(x = reorder(Predictor, -Importance), y = Importance)) +
  geom_bar(stat = "identity", fill = "blue") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(x = "Predictor", y = "Average Importance Score") +
  ggtitle("Overall Predictor Importance Across All Models")

```
