---
title: "Untitled"
author: "Mamadou"
date: "2023-11-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Initial Setup: Required Libraries and Working Directory
```{r message=F}
library(tidyverse)
library(sf)
library(plotly)
library(caret)
library(likert)
library(grid)
library(gridExtra)
library(ggpubr)
```


## Data Wrangling

## Load and Inspect the Data

```{r}
# load the data
data <- read.csv("diabetes_health_indicators.csv")
```

Removing unwanted variables

```{r}
data_select <- data %>%
        select(-c(3,7,14,18,))
```

let's check the structure of the dataset
```{r}
str(data_select)
```
We can see that all data are numeric. However some are actually categorical ones so we need to convert them to factor data type.

```{r}
data_select$Diabetes_binary <- as.factor(data_select$Diabetes_binary)
data_select$HighBP <- as.factor(data_select$HighBP)
data_select$CholCheck <- as.factor(data_select$CholCheck)
data_select$Smoker <- as.factor(data_select$Smoker)
data_select$HeartDiseaseorAttack <- as.factor(data_select$HeartDiseaseorAttack)
data_select$PhysActivity <- as.factor(data_select$PhysActivity)
data_select$Fruits <- as.factor(data_select$Fruits)
data_select$Veggies <- as.factor(data_select$Veggies)
data_select$HvyAlcoholConsump <- as.factor(data_select$HvyAlcoholConsump)
data_select$AnyHealthcare <- as.factor(data_select$AnyHealthcare)
data_select$Sex <- as.factor(data_select$Sex)
```

## Data Subset
Since we have 70692 observations, we need to subset the data a lower dataset and we opt to move to 2000 observation and 9 variables.
```{r}
data_subset <- data_select[1:2000, ]
```